---
title: "Qslice package"
author: "Matthew J. Heiner, David B. Dahl, and Samuel B. Johnson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Qslice package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library("qslice")
```

The `qslice` package implements the quantile slice sampler of Heiner et al., (2024+) as well as other popular slice samplers. It also includes several utility functions for specifying psueudo-target distributions, for diagnostics, and for tuning.

Other implemented methods include the generalized elliptical (Nishihara et al., 2014), latent (Li and Walker, 2023), and stepping-out (Neal, 2003) slice samplers, and independence Metropolis-Hastings.

All sampling functions in `qslice` perform a single scan that can be iteratively called as part of an MCMC routine.

# Slice sampling

Let $g(x)$ represent an unnormalized target density. Augment the target with a latent random variable $(V \mid X = x) \sim \text{Uniform}(0, g(x))$, yielding a joint density for $X$ and $V$ that is constant.

Simple slice samplers first draw from the conditional distribution of $V$ given above, and then from the conditional of $X \mid V = v$, which is uniform on the set $A = \{ x : v < g(x) \}$ known as the slice region. If $A$ is available analytically, a custom slice sampler can be implemented. The slice samplers available in this package repeatedly sample until a draw on the slice region is found. All use the shrinking method in Neal (2003), which we demonstrate below.

A Gibbs sampler drawing $V_t \mid X_{t-1}$ followed by $X_t \mid V_t$ will produce a sequence of draws $\{X_t\}$ whose distribution converges to the marginal density proportional to $g(x)$.

## Example

Each sampling method requires a function to evaluate the natural logarithm of a (typically unnormalized) target density. We will use a target that is proportional to a gamma density
$$ g(x) = x^{\alpha - 1} \exp(-x) \, 1_{(x>0)} \, .$$

```{r target}
## use gamma(shape = alpha) as a target
alpha <- 2.5
ltarget <- function(x) ifelse(x > 0.0, (alpha - 1.0) * log(x) - x, -Inf)
```

Note that log-target functions in `qslice` rely on lexical scoping. That is, parameter `alpha` is NOT passed as an argument to function `ltarget`. If the target distribution changes at each iteration of MCMC, the function will need to be redefined.

We will first demonstrate the step-and-shrink sampler of Neal (2003), an excellent general tool that is robust to the choice of its tuning parameter `w`. 
The sampling function `slice_stepping_out` performs a single draw, so we embed it within an MCMC algorithm.

```{r step}
## set up MCMC
n_iter <- 1e3
x_sample <- numeric(n_iter + 1)
x_sample[1] <- 0.5  # initialize
n_eval <- 0  # count target evaluations

## run step-and-shrink slice sampler
for (i in 2:(n_iter+1)) {
  state <- slice_stepping_out(x = x_sample[i-1], 
                              log_target = ltarget,
                              w = 2.0)
  n_eval <- n_eval + state$nEvaluations
  x_sample[i] <- state$x
}

## check samples
n_eval / n_iter  # target evaluations per iteration of MCMC
hist(x_sample, freq = FALSE, n = 30)
curve(dgamma(x, shape = alpha), col = "blue", lwd = 2, add = TRUE)

## null hypothesis: x_sample ~iid gamma(alpha)
ks.test(x_sample[seq(1, n_iter, by = 5)], pgamma, shape = alpha)
```


# Quantile slice sampler

The quantile slice sampler uses an approximation to the target density, called a pseudo-target. Let $\hat{G}(\cdot)$ represent the distribution function (CDF) of the pseudo-target with inverse (quantile) function $\hat{G}^{-1}(\cdot)$ and density $\hat{g}(\cdot)$. We use the probability integral transform to define a new random variable $\psi = \hat{G}(X) \in [0,1]$. The original target can be written as
$$g(x) = \frac{g(x)}{\hat{g}(x)}\hat{g}(x) \, ,$$
which after transformation becomes
$$h(\psi) = \frac{g(\hat{G}^{-1}(\psi))}{\hat{g}(\hat{G}^{-1}(\psi))} \, 1_{(0 \le \psi \le 1)}  \, .$$
If the pseudo-target approximates the target well, $h(\psi)$ will be close to a constant function, yielding an efficient slice sampler. The quantile slice sampler uses target $h(\psi)$ and adaptively shrinks (around the previously sampled value of $\psi$) to draw a sample belonging the slice region $A_\psi \subseteq [0,1]$.


## Example

The quantile slice sampler requires a pseudo-target. This is specified in `qslice` with a list containing two functions: the log-density `ld` and inverse-CDF (quantile) `q` functions. `qslice` offers pseudo-target constructor functions with options for several distribution families. See `help(pseudo_list)`. Here we will use a truncated $t$ distribution:

```{r pseudo1}
pseu <- pseudo_list(family = "t", 
                    params = list(loc = 0, sc = 3, degf = 1), 
                    lb = 0)
```

When specifying a pseudo-target, remember: 

1. The support of the pseudo-target should match that of the original target (note the use of `lb = 0` to set the pseudo-target's lower bound in the call to `pseudo_list`).
2. The pseudo-target should have tails that are at least as heavy as the original target (we recommend Student-$t$ for this reason).

We can visualize $g(x)$, $\hat{g}(x)$ and $h(\psi)$.

```{r vis_pseudo1, fig.width=4.5, fig.height=3}
utility_pseudo(pseudo = pseu, log_target = ltarget, 
               type = "function", 
               utility_type = "AUC", plot = TRUE)
```

The function `utility_pseudo` measures how close $h(\cdot)$ is to a constant function. Here we use AUC $\in (0, 1]$, defined as the area under the curve $h(\psi) / \max_z(h(z))$.

The function `slice_quantile` performs a single draw using the quantile slice sampler. It also outputs draws of $\psi$ (called `u` in the output), which can be used as a diagnostic for pseudo-target fitness.

```{r qslice1, fig.show='hold'}
## set up MCMC
n_iter <- 1e3
x_sample <- psi_sample <- numeric(n_iter + 1)
x_sample[1] <- psi_sample[1] <- 0.5  # initialize
n_eval <- 0  # count target evaluations

## run step-and-shrink slice sampler
for (i in 2:(n_iter+1)) {
  state <- slice_quantile(x = x_sample[i-1], 
                          log_target = ltarget,
                          pseudo = pseu)
  n_eval <- n_eval + state$nEvaluations
  x_sample[i] <- state$x
  psi_sample[i] <- state$u
}

## check samples
n_eval / n_iter  # target evaluations per iteration of MCMC

hist(x_sample, freq = FALSE, n = 30)
curve(dgamma(x, shape = alpha), col = "blue", lwd = 2, add = TRUE)

hist(psi_sample, freq = FALSE, n = 30)

auc(u = psi_sample) # calculate AUC from transformed samples

## null hypothesis: x_sample ~iid gamma(alpha)
ks.test(x_sample[seq(1, n_iter, by = 10)], pgamma, shape = alpha)
```

## Optimal pseudo-targets and tuning

The quantile slice sampler is effective even with crude approximations to the target. We can also find a pseudo-target within the Student-$t$ family that maximizes AUC.

```{r pseudo2, fig.width=4.5, fig.height=3}
pseu_opt <- pseudo_opt(log_target = ltarget, 
                       type = "function",
                       family = "t", degf = c(1, 5, 20), 
                       lb = 0,
                       utility_type = "AUC", plot = TRUE)
```

Alternatively, we can use initial samples from the target to specify a pseudo-target. This gives a simple procedure for "tuning" the sampler.

```{r pseudo3}
pseu_opt <- pseudo_opt(samples = x_sample, 
                       type = "samples",
                       family = "t", degf = c(1, 5, 20), 
                       lb = 0,
                       utility_type = "AUC", plot = TRUE, nbins = 20)

names(pseu_opt)
names(pseu_opt$pseudo)
pseu_opt$pseudo$txt
```


We conclude by running the quantile slice sampler again with an optimized pseudo-target.

```{r qslice2, fig.show='hold'}
## set up MCMC
n_iter <- 1e3
x_sample <- psi_sample <- numeric(n_iter + 1)
x_sample[1] <- psi_sample[1] <- 0.5  # initialize
n_eval <- 0  # count target evaluations

## run step-and-shrink slice sampler
for (i in 2:(n_iter+1)) {
  state <- slice_quantile(x = x_sample[i-1], 
                          log_target = ltarget,
                          pseudo = pseu_opt$pseudo)
  n_eval <- n_eval + state$nEvaluations
  x_sample[i] <- state$x
  psi_sample[i] <- state$u
}

## check samples
n_eval / n_iter  # target evaluations per iteration of MCMC

hist(x_sample, freq = FALSE, n = 30)
curve(dgamma(x, shape = alpha), col = "blue", lwd = 2, add = TRUE)

hist(psi_sample, freq = FALSE, n = 30)

auc(u = psi_sample) # calculate AUC from transformed samples

## null hypothesis: x_sample ~iid gamma(alpha)
ks.test(x_sample[seq(1, n_iter, by = 10)], pgamma, shape = alpha)
```

# References

- Heiner, M. J., Johnson, S. B., Christensen, J. R., and Dahl, D. B. (2024+), "Quantile Slice Sampling," *arXiv preprint arXiv:###*. 

- Li, Y. and Walker, S. G. (2023), "A latent slice sampling algorithm," *Computational Statistics and Data Analysis*, 179, 107652. <https://doi.org/10.1016/j.csda.2022.107652>

- Neal, R. M. (2003), "Slice sampling," *The Annals of Statistics*, 31, 705-767. <https://doi.org/10.1214/aos/1056562461>

- Nishihara, R., Murray, I., and Adams, R. P. (2014), "Parallel MCMC with Generalized Elliptical Slice Sampling," *Journal of Machine Learning Research*, 15, 2087-2112. <https://jmlr.org/papers/v15/nishihara14a.html>
